### 2017-10-25 Paper_v6.tex
adding updated manuscript

Villaverde and co-workers published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
They ranked the example problems by computational cost from most to least expensive.
We evaluated the performance of DOPS on problems from the least and most expensive categories.
The least expensive problem (henceforth referred to as problem B4) was a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters \cite{villaverde2014high}.
The biochemical reactions were modeled using modular rate laws and generalized Michaelis-Menten kinetics.
On the other hand, the expensive problem was a genome scale kinetic model of \textit{Saccharomyces~cerevisiae} with 261 reactions, 262 variables and 1759 parameters \cite{smallbone2013large} (henceforth referred to as problem B1).
In both cases, synthetic time series data generated with known parameter values, was used to estimate model parameters.
For problem B1, the time series data consisted of 44 observables, and for problem B4 the data corresponded to 13 different metabolite measurement sets.
We fixed the number of function evaluations at $\mathcal{N}$ = 4000, and trained both models against the synthetic experimental data.
DOPS produced good fits to the synthetic data (Fig. \ref{fig-sims-b1} and Fig. \ref{fig-sims-b4}), and recapitulated the nominal parameter values using only $\mathcal{N}\leq~4000$ function evaluations (Fig. \ref{fig-benchmark}).
On the other hand, enhanced scatter search (eSS) with a local optimizer took on order $10^{5}$ function evaluations for the same problems.
Thus, DOPS estimated the parameters in benchmark biochemical models, and recovered the original parameters from synthetic data, using fewer function evaluations.
Next, we compared the performance of DOPS with the four other meta-heuristics for a model of the human coagulation cascade.

@@

\subsubsection*{Phase 1: Particle swarm phase.}
The particle swarm phase of DOPS begins by randomly initializing a swarm of $\mathcal{K}$-dimensional particles (represented as $z_{i}$),
wherein each particle corresponded to a $\mathcal{K}$-dimensional parameter vector.
After initialization, particles were randomly partitioned into $k$ equally sized sub-swarms $\mathcal{S}_{1},\hdots,\mathcal{S}_{k}$.
Particles within each sub-swarm $\mathcal{S}_{k}$ were updated according to the rule:
\begin{equation}\label{eqn:update-rule}
	%\mathbf{z}_{i,j} = \theta_{1,j-1}\mathbf{z}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{z}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G}_{k} - \mathbf{z}_{i,j-1}\right)
	{z}_{i,j} = \theta_{1,j-1}{z}_{i,j-1} + \theta_{2}{r}_{1}\left(\mathcal{L}_{i} - {z}_{i,j-1}\right) + \theta_{3}{r}_{2}\left(\mathcal{G}_{k} - {z}_{i,j-1}\right)
\end{equation}
where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ were adjustable parameters, $\mathcal{L}_{i}$ denotes the best solution found by particle $i$ within sub-swarm
$\mathcal{S}_{k}$ for function evaluation $1\rightarrow j-1$, and
$\mathcal{G}_{k}$ denotes the best solution found over all particles within sub-swarm $\mathcal{S}_{k}$.
The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model parameters ($\mathcal{K}\times{1}$).
Equation \eqref{eqn:update-rule} is similar to the general particle swarm update rule, however, it does not contain velocity terms.
In DOPS, the parameter $\theta_{1,j-1}$ is similar to the inertia weight parameter for the velocity term described by Shi and Eberhart \cite{shi1999empirical};
Shi and Eberhart proposed a linearly decreasing inertia weight to improve convergence properties of particle swarm optimization.
Our implementation of $\theta_{1,j-1}$ is inspired by this and the decreasing perturbation probability proposed by Tolson and Shoemaker \cite{tolson2007dynamically}.
It is an analogous equivalent to inertia weight on velocity.
However $\theta_{1,j-1}$ places inertia on the position rather than velocity and uses the same rule described by Shi and Eberhart to adaptively change with the number of function evaluations:
\begin{eqnarray}
	\mathbf \theta_{1,j}&=&\frac{(\mathcal{N}-{j})*({w}_{max}-{w}_{min}))}{(\mathcal{N}-{1})} + {w}_{min}
\end{eqnarray}
where $\mathcal{N}$ represents the total number of function evaluations, ${w}_{max}$ and ${w}_{min}$ are the maximum and minimum inertia weights, respectively.
While updating the particles, parameter bounds were enforced using reflection boundary conditions (Algo. \ref{algo_reflection}).

@@

\begin{figure}%
    \centering
    \subfloat[]{{\includegraphics[trim = {3cm 7cm 3cm 7cm }, clip,width=.45\textwidth]{./rachelfigs/FigureS7a} }}%
    \qquad
    \subfloat[]{{\includegraphics[trim = {3cm 7cm 3cm 7cm }, clip,width=.45\textwidth]{./rachelfigs/FigureS7b} }}%
    \qquad
        \subfloat[]{{\includegraphics[trim = {3cm 7cm 3cm 7cm }, clip,width=.45\textwidth]{./rachelfigs/FigureS7c} }}%
    \qquad
    \subfloat[]{{\includegraphics[trim = {3cm 7cm 3cm 7cm }, clip,width=.45\textwidth]{./rachelfigs/FigureS7d} }}%
    \caption{Variability  analysis in best objective value for $\mathcal{T}$ = 25 trials. (a) Ackley  300 dimensional (b) Rastrigin 300 dimensional (c) CHO model (problem b4) (d) coagulation. }%
    \label{fig-DOPS-Dispersion-Curves}%
\end{figure}

>>>>

Villaverde and co-workers published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
They ranked the example problems by computational cost from most to least expensive.
We evaluated the performance of DOPS on problems from the least and most expensive categories.
The least expensive problem (henceforth referred to as problem B4) was a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters \cite{villaverde2014high}.
The biochemical reactions were modeled using modular rate laws and generalized Michaelis-Menten kinetics.
On the other hand, the expensive problem was a genome scale kinetic model of \textit{Saccharomyces~cerevisiae} with 261 reactions, 262 variables and 1759 parameters \cite{smallbone2013large} (henceforth referred to as problem B1).
In both cases, synthetic time series data generated with known parameter values, was used to estimate model parameters.
For problem B1, the time series data consisted of 44 observables, and for problem B4 the data corresponded to 13 different metabolite measurement sets.
We fixed the number of function evaluations at $\mathcal{N}$ = 4000, and trained both models against the synthetic experimental data.
DOPS produced good fits to the synthetic data (Fig. \ref{fig-sims-b1} and Fig. \ref{fig-sims-b4}), and recapitulated the nominal parameter values using only $\mathcal{N}\leq~4000$ function evaluations (Fig. \ref{fig-benchmark}).
On the other hand, enhanced scatter search (eSS) with a local optimizer took on order $10^{5}$ function evaluations for the same problems. DOPS also had lower variability in the best value obtained (Fig \ref{fig-DOPS-Variability}) and faster convergence (Fig \ref{fig-more-methods}, Fig \ref{fig-compare-DOPS-ESS}) across multiple runs when compared to other metaheuristics while requiring a comparable amount of time (Fig \ref{fig-time-comparsion}). 
Thus, DOPS estimated the parameters in benchmark biochemical models, and recovered the original parameters from synthetic data, using fewer function evaluations.
Next, we compared the performance of DOPS with the four other meta-heuristics for a model of the human coagulation cascade.

@@

\subsubsection*{Phase 1: Particle swarm phase.}
Particle warm optimization is an evolutionary algorithm that uses a population of particles (solutions) to find an optimal solution. Each particle is updated based on its experience (particle best) and the experience of all other particles within the swarm (global best). 
The particle swarm phase of DOPS begins by randomly initializing a swarm of $\mathcal{K}$-dimensional particles (represented as $z_{i}$),
wherein each particle corresponded to a $\mathcal{K}$-dimensional parameter vector.
After initialization, particles were randomly partitioned into $k$ equally sized sub-swarms $\mathcal{S}_{1},\hdots,\mathcal{S}_{k}$.
Particles within each sub-swarm $\mathcal{S}_{k}$ were updated according to the rule:
\begin{equation}\label{eqn:update-rule}
	%\mathbf{z}_{i,j} = \theta_{1,j-1}\mathbf{z}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{z}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G}_{k} - \mathbf{z}_{i,j-1}\right)
	{z}_{i,j} = \theta_{1,j-1}{z}_{i,j-1} + \theta_{2}{r}_{1}\left(\mathcal{L}_{i} - {z}_{i,j-1}\right) + \theta_{3}{r}_{2}\left(\mathcal{G}_{k} - {z}_{i,j-1}\right)
\end{equation}
where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ were adjustable parameters, $\mathcal{L}_{i}$ denotes the best solution found by particle $i$ within sub-swarm
$\mathcal{S}_{k}$ for function evaluation $1\rightarrow j-1$, and
$\mathcal{G}_{k}$ denotes the best solution found over all particles within sub-swarm $\mathcal{S}_{k}$.
The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model parameters ($\mathcal{K}\times{1}$).
Equation \eqref{eqn:update-rule} is similar to the general particle swarm update rule, however, it does not contain velocity terms.
In DOPS, the parameter $\theta_{1,j-1}$ is similar to the inertia weight parameter for the velocity term described by Shi and Eberhart \cite{shi1999empirical};
Shi and Eberhart proposed a linearly decreasing inertia weight to improve convergence properties of particle swarm optimization.
Our implementation of $\theta_{1,j-1}$ is inspired by this and the decreasing perturbation probability proposed by Tolson and Shoemaker \cite{tolson2007dynamically}.
It is an analogous equivalent to inertia weight on velocity.
However $\theta_{1,j-1}$ places inertia on the position rather than velocity and uses the same rule described by Shi and Eberhart to adaptively change with the number of function evaluations:
\begin{eqnarray}
	\mathbf \theta_{1,j}&=&\frac{(\mathcal{N}-{j})*({w}_{max}-{w}_{min}))}{(\mathcal{N}-{1})} + {w}_{min}
\end{eqnarray}
where $\mathcal{N}$ represents the total number of function evaluations, ${w}_{max}$ and ${w}_{min}$ are the maximum and minimum inertia weights, respectively.
While updating the particles, parameter bounds were enforced using reflection boundary conditions (Algo. \ref{algo_reflection}).

@@

\begin{figure}%
    \centering
    \subfloat[]{{\includegraphics[trim = {3cm 7cm 3cm 7cm }, clip,width=.45\textwidth]{./rachelfigs/FigureS7a} }}%
    \qquad
    \subfloat[]{{\includegraphics[trim = {3cm 7cm 3cm 7cm }, clip,width=.45\textwidth]{./rachelfigs/FigureS7b} }}%
    \qquad
        \subfloat[]{{\includegraphics[trim = {3cm 7cm 3cm 7cm }, clip,width=.45\textwidth]{./rachelfigs/FigureS7c} }}%
    \qquad
    \subfloat[]{{\includegraphics[trim = {3cm 7cm 3cm 7cm }, clip,width=.45\textwidth]{./rachelfigs/FigureS7d} }}%
    \caption{Variability  analysis in best objective value for $\mathcal{T}$ = 25 trials. (a) Ackley  300 dimensional (b) Rastrigin 300 dimensional (c) CHO model (problem b4) (d) coagulation. }%
    \label{fig-DOPS-Variability}%
\end{figure}
